A primer to numerical simulations: The perihelion motion of Mercury

### 2018-02-05 tex/MercuryNumerical.tex
Added addition why no terms linear in rL plus additional ref

\author{
	C.~K\"orber$^{1}$,
	I.~Hammer$^{1}$,
	J.-L.~Wynen$^{1}$,
	J.~Heuer$^{2}$,
	C.~M\"uller$^{3}$ and
	C.~Hanhart$^{1}$
}
\address{
	$^1$ \textit{Institut f\"ur Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
	$^2$ \textit{Institut f\"ur Neurowissenschaften und Medizin (INM-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
	$^3$ \textit{JuLab, Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}
}
\ead{c.koerber@fz-juelich.de, c.hanhart@fz-juelich.de}
\vspace{10pt}
%\begin{indented}
%\item[]February 2014
%\end{indented}

@@

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha \, \frac{r_S}{r} )$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from a $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_L^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_L^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta \, \frac{r_L^2}{r^2}$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_L^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_L^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{ r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$~km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other (see Figure \ref{fig:sun_merc}).
We thus find that $(r_L^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
Therefore one may expect from both correction terms an effect of similar size --- for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}.
Furthermore, any term of higher order in either $r_L^2$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General Relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

@@

Dimensional analysis is not only a tool that allows one to cross check, if the results of simulations are of the right order of magnitude, it is also very helpful to identify unusual dynamics in some systems.
Especially the latter aspect should become clear from the discussion in this section.

The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities (like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order one (that means anything between about 0.1 and 10 is fine --- but 0.01 or 100 is irritating); parameters in line with this are called "natural".
Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would predict the parameters $\alpha$ and $\beta$ to be
of order one.
One expects that unnatural parameters point at dynamics not accounted for explicitly.
Employing for the problem at hand the average distance Mercury-Sun $\bar r_{MS}=6\times 10^7\,\mathrm{km}$, the concept on 
naturalness allows us to estimate the expected angular shift per orbit for, e.g., $\alpha=1$ and $\beta=0$
\begin{equation}
\delta \Theta \simeq 2\pi\left(\frac{r_S}{\bar r_{MS}}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\cdot 10^{-5})^o = (7\cdot 10^{-2}) ^{''} \ ,
\end{equation}
or for $\alpha=0$ and $\beta=1$
\begin{equation}
\delta \Theta \simeq 2\pi\left(\frac{a^2}{\bar r^2_{MS}}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (1 \cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
\end{equation}
which leads to a shift of about $30^{''}$ and $15^{''}$, respectively, in 100 earth years. These are to be compared to the empirical value of $43^{''}$.
Thus the amount of perihelion motion of Mercury is indeed in line with expectations, \textit{if} --- and this is an important "if" --- the 
Newtonian dynamics is simply the leading term of some more general underlying theory. 
 In particular, no new scales enter in the correction terms in addition to $r_S$ and $r_L^2$. This is by itself already an
 interesting observation.
 In case of General Relativity $\beta=3$ is a natural value 
while $\alpha$ vanishes. This pattern is therefore a non-trivial prediction of General Relativity and one should expect
that alternative theories of gravitation generate non-vanishing values of $\alpha$.

>>>>

\author{
	C.~K\"orber$^{1}$,
	I.~Hammer$^{1}$,
	J.-L.~Wynen$^{1}$,
	J.~Heuer$^{2}$\footnote{Present Address: Institut f\"ur Neurowissenschaften und Medizin (INM-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany},
	C.~M\"uller$^{3}$ and
	C.~Hanhart$^{1}$
}
\address{
	$^1$ \textit{Institut f\"ur Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
	$^2$ \textit{Hochschule Hamm-Lippstadt, Marker Allee 76-78, 59063 Hamm, Germany}\\
	$^3$ \textit{Sch\"ulerlabor JuLab, Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}
}
\ead{c.koerber@fz-juelich.de, c.hanhart@fz-juelich.de}
\vspace{10pt}
%\begin{indented}
%\item[]February 2014
%\end{indented}

@@

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha \, \frac{r_S}{r} )$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from a $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_L^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_L^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta \, \frac{r_L^2}{r^2}$.
Here one might wonder why this additional term is chosen proportional to $(r_L/r)^2$ and not to $(r_L/r)$. The reason for this choice
is indeed not obvious: In general the correction terms added must be scalar quantities --- accordingly vectors can enter only
as scalar products. Moreover, one is not allowed to use square-roots of those scalar products for these impose
wrong mathematical properties to the equations of motion (for the case at hand, e.g., the derivative with
respect to the velocity would not be defined at $\dot{\vec r}=0$) --- the only vector that is allowed to appear as its length in linear order
is the radius, since this length is related to geometrical properties of the system.  
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_L^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_L^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{ r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$~km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other (see Figure \ref{fig:sun_merc}).
We thus find that $(r_L^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
Therefore one may expect from both correction terms an effect of similar size --- for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}.
Furthermore, any term of higher order in either $r_L^2$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General Relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).
Modifications to the Newtonian equation of gravity introduced to account for the perihelion motion of Mercury
are discussed in great detail in a very pedagogical way also in Ref.~\cite{Wells:2011st}.

@@

Dimensional analysis is not only a tool that allows one to cross check, if the results of simulations are of the right order of magnitude, it is also very helpful to identify unusual dynamics in some systems.
Especially the latter aspect should become clear from the discussion in this section. The modifications to the Newtonian equation of gravity introduced to account for the perihelion motion of Mercury
are discussed also in Ref.~\cite{Wells:2011st}. To get a deeper understanding of the concepts introduced in this section reading this article is highly recommended.

The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities (like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order one (that means anything between about 0.1 and 10 is fine --- but 0.01 or 100 is irritating); parameters in line with this are called "natural".
Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would predict the parameters $\alpha$ and $\beta$ to be
of order one.
One expects that unnatural parameters point at dynamics not accounted for explicitly.
Employing for the problem at hand the average distance Mercury-Sun $\bar r_{MS}=6\times 10^7\,\mathrm{km}$, the concept on 
naturalness allows us to estimate the expected angular shift per orbit for, e.g., $\alpha=1$ and $\beta=0$
\begin{equation}
\delta \Theta \simeq 2\pi\left(\frac{r_S}{\bar r_{MS}}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\cdot 10^{-5})^o = (7\cdot 10^{-2}) ^{''} \ ,
\end{equation}
or for $\alpha=0$ and $\beta=1$
\begin{equation}
\delta \Theta \simeq 2\pi\left(\frac{r_L^2}{\bar r^2_{MS}}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (1 \cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
\end{equation}
which leads to a shift of about $30^{''}$ and $15^{''}$, respectively, in 100 earth years. These are to be compared to the empirical value of $43^{''}$.
Thus the amount of perihelion motion of Mercury is indeed in line with expectations, \textit{if} --- and this is an important "if" --- the 
Newtonian dynamics is simply the leading term of some more general underlying theory. 
 In particular, no new scales enter in the correction terms in addition to $r_S$ and $r_L^2$. This is by itself already an
 interesting observation.
 In case of General Relativity $\beta=3$ is a natural value 
while $\alpha$ vanishes. This pattern is therefore a non-trivial prediction of General Relativity and one should expect
that alternative theories of gravitation generate non-vanishing values of $\alpha$.
